A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-13 sections/Overview.tex
revise 2.2 + comment

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Broken Metatheory} % Talk about the lemmas that are false, show counterexample.
While going through the manual proofs of DK's algorithm, we found several problems.
The proof of Lemma 19 (Extension Equality Preservation) applies a wrong
induction hypothesis for the ${\longrightarrow}\mathtt{Uvar}$ case.
Fortunately the lemma could be proven without the well-formedness conditions
$\Gm\vdash A$ and $\Gm\vdash B$.
Similar problem happens for Lemma 14 (Subsumption), where many applications of
induction hypotheses are not correct.
For manual proofs, induction hypotheses are not automatically generated by programs,
thus are easily misused.

Moreover, Lemma 29 (Parallel Admissibility) is false.
One may find a counterexample with detailed explaination in the appendix.
Lemma 29 is an important lemma for the proof of soundness and completeness,
which are two main results of their metatheory.
Instantiation soundness (i.e. a part of subtyping soundness) and typing completeness
both directly require Lemma 29.
Lemma 54 (Typing Extension) requires Lemma 29 and is applied 13 times
in the proof of typing soundness and completeness.

@@

DK is already quite disciplined on these accounts, in particular compared to
traditional constraint-based type-inference algorithms like Algorithm W which
features an implicit global scope for all type variables. Indeed, DK uses an
explicit and ordered context that tracks the relative scope of universal and
existential variables and it is careful to only instantiate existential
variables in a well-scoped manner.

Moreover, DK's algorithm carefully propagates instantiations by recording them
into the context and threading this context through all judgments.  means of
two This means that every judgments takes two contexts---an input and an output
context---rather than the conventional single context. The output context
records any new variable instantations; to propagate these instantiations to
the remaining judgements, their predecessor's output context---which is their
input context---is applied to them as a substitution.

While it works, DK's approach is still fairly involved and thus hard to reason
about in a mechanised setting. Indeed, the instantiations have to be recorded
in the context and are applied incrementally to each remaining judgment in
turn, rather than immediately to all remaining judgments at once. Also the
stack discipline of the ordered contexts does not mesh well with the use of
output contexts; explicit marker entries are needed in two places to
demarcate the end of an existential variable's scope. These are compelling 
reasons to look for a simpler algorithm that is easier to reason about in a 
mechanised setting.

>>>>

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Broken Metatheory} % Talk about the lemmas that are false, show counterexample.
While going through the manual proofs of DK's algorithm, we found several
problems.  Indeed, two proofs of lemmas---Lemma 19 (Extension Equality
Preservation) and Lemma 14 (Subsumption)--- wrongly apply induction hypotheses
in several cases. Fortunately, we have found simple work-arounds that fix these
proofs without affecting the appeals to these lemmas.

% The proof of Lemma 19 (Extension Equality Preservation) applies a wrong
% induction hypothesis for the ${\longrightarrow}\mathtt{Uvar}$ case.
% Fortunately the lemma could be proven without the well-formedness conditions
% $\Gm\vdash A$ and $\Gm\vdash B$.
% Similar problem happens for Lemma 14 (Subsumption), where many applications of
% induction hypotheses are not correct.
% For manual proofs, induction hypotheses are not automatically generated by programs,
% thus are easily misused.

More seriously, we have also found a lemma that simply does not hold, 
Lemma 29 (Parallel Admissibility). See Appendix~\ref{} for a detailed explanation
and counterexample. This lemma is a cornerstone of the two metatheoretical results 
of the algorithm, soundness and completeness with respect to the declarative system.
In particular, both instantiation soundness (i.e. a part of subtyping
soundness) and typing completeness both directly require the broken lemma.
Moreover, Lemma 54 (Typing Extension) also requires the broken lemma and is
itself used 13 times in the proof of typing soundness and completeness.
Unfortunately, we have not yet found a way to fix this problem.
\tom{You have communicated this to the authors and not received a fix, right?
     We could mention that too then.}

@@

DK is already quite disciplined on these accounts, in particular compared to
traditional constraint-based type-inference algorithms like Algorithm W which
features an implicit global scope for all type variables. Indeed, DK uses its
explicit and ordered context $\Gamma$ that tracks the relative scope of universal and
existential variables and it is careful to only instantiate existential
variables in a well-scoped manner.

Moreover, DK's algorithm carefully propagates instantiations by recording them
into the context and threading this context through all judgments. 
% two This means that every judgments takes two contexts---an input and an output
% context---rather than the conventional single context. The output context
% records any new variable instantations; to propagate these instantiations to
% the remaining judgements, their predecessor's output context---which is their
% input context---is applied to them as a substitution.
While this works well on paper, this approach is still fairly involved and thus
hard to reason about in a mechanised setting. Indeed, the instantiations have
to be recorded in the context and are applied incrementally to each remaining
judgment in turn, rather than immediately to all remaining judgments at once.
Also, as we have mentioned above, the stack discipline of the ordered contexts
does not mesh well with the use of output contexts; explicit marker entries are
needed in two places to demarcate the end of an existential variable's scope.
These are compelling reasons to look for a simpler algorithm that is easier to
reason about in a mechanised setting.
